\emph{Specific taxa differentiate proximal and distal microbiota in the unprepped human colon}

### 2017-06-19 submission/manuscript.tex
final tweaks

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={**Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota*},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{**Machine learning identifies unique taxa differentiating proximal and
distal human colonic microbiota*}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide.
Tumors of the proximal (right) and distal (left) colon are
morphologically and genetically distinct. Previous work from our group
found that microbial dysbiosis is associated with the development of CRC
tumors in studies of both mice and humans. Analysis of the fecal
microbiota from healthy and CRC patients further revealed different
microbial signatures associated with disease. In this study, we extended
our observations of the fecal microbiome to analysis of the proximal and
distal human colon. We used a two-colonoscope approach on subjects that
had not undergone standard bowel preparation procedure. This technique
allowed us to characterize the native proximal and distal luminal and
mucosal microbiome without prior chemical disruption. 16S rRNA gene
sequencing was performed on proximal and distal mucosal and luminal
biopsies and exit stool for 20 healthy individuals. Diversity analysis
revealed that each site contained a diverse community, and that a
patient's samples were more similar to each other than to that of other
individuals. Since we could not differentiate sites along the colon
based on community structure or community membership alone, we employed
the machine-learning algorithm Random Forest to identify key species
that distinguish biogeographical sites. Random Forest classification
models were built using taxa abundance and sample location and revealed
distinct populations that were found in each location.
\emph{Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas} and
\emph{Actinomyces} were most likely to be found in mucosal samples
versus luminal samples (AUC = 0.925). The classification model performed
well (AUC = 0.912) when classifying mucosal samples into proximal or
distal sides, but separating luminal samples from each side proved more
challenging (AUC = 0.755). The left mucosa was found to have high
populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
Proximal and distal luminal samples were comprised of many of the same
taxa, likely reflecting the fact that stool moves along the colon from
the proximal to distal end. Finally, comparison of all samples to fecal
samples taken at exit uncovered that the feces were most similar to
samples taken from the left lumen, again reflecting the anatomical
structure of the colon. Taken together, our results have identified
distinct bacterial populations of the proximal and distal colon. Further
investigation of these bacteria may elucidate if and how these groups
contribute to differential oncogenesis processes on the respective sides
of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related
deaths in the United States. CRC tumors vary in structure, size,
morbidity and symptomology depending upon their geographical location in
the colon. CRC has recently been described to fall into four molecular
subtypes ((1)). Of these subtypes, tumors that arise on the left side of
the colon are infiltrating lesions that present with painful symptoms.
In contrast, right-sided colon tumors are bulky and project into the
lumen, often causing no symptoms until they have developed into
carcinomas ((2)). Because of the absence of symptoms, right-sided
tumors, which comprise up to 47\% of CRCs, have a significantly higher
mortality rate ((2)). The left and right sides of the colon differ in
the amount of inflammation present and the genomic instability of
precancerous cells, respectively, as well as oxygen, pH and the presence
of antimicrobial peptides ((3), (4), (5)). Microenvironments differ not
only longitutinally along the colon, but latitudinally from the
epithelium to mucosa to intestinal lumen, offering several sites for
different microbial communities to flourish. Given this varied
physiology of the proximal-distal axis of the colon, symbiotic microbes
and their metabolites likely vary between sites.

Several recent findings have shown that development and progression of
CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host ((3)). Our group
and others have found that the stool microbiome of patients with CRC is
distinct from that of healthy people ((6)). Further studies manipulating
the gut microbiome using antibiotics or other chemoagents in mice has
shown that dysbiosis preceded and accelerated the development of CRC
tumors ((7)). Comparison of the specific bacteria present on CRC tumors
with those found on nearby healthy tissue has also identified specific
bacterial species that are tumor-associated based on FISH staining and
16S rRNA sequencing ((8)). These species include the oral pathogens
\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
Interestingly, the periodontal pathogens are also highly predictive of
if a patient has CRC tumors or not in our classification studies ((9)).
However, these studies have been performed comparing shed stool or in
mice, which does not allow for analysis of paried samples from the
proximal and distal sides of the colon. Similarly, comparisons of on- or
off-tumor bacteria rarely have matched tissue from the other side of the
colon from the same patient, limiting what conclusions can be drawn
about the colonic microbiome overall, let alone at that specific site.
Disruption of the healthy community could promote the initiation or
proliferation of the distinct left and right CRC tumors. However, the
contribution of the gut microbiota to these subtypes is largely
undefined.

Furthermore, profiles of the microbial biogeography of the gut have been
limited by sample collection methods. The majority of human gut
microbiome studies have been performed on whole shed stool or on samples
collected during colonoscopy procedures. While the latter method allows
investigators to acquire samples from inside the human colon, typically
this procedure is preceded by the use of bowel preparation methods such
as the consumption of laxatives to cleanse the bowel. Bowel preparation
is essential for detecting cancerous or precancerous lesions in the
colon, but complicates microbiome profiling as the chemicals strip the
bowel of contents and disrupt the mucosal layer ((10), (11)). As such,
what little information we do have about the biogeographical
distribution of the microbes in the proximal and distal colon is
confounded by the bowel preparation procedure.

In this study we sought to combine these approaches to characterize the
microbiome in the lumen and mucosa of the proximal and distal healthy
human colon. Our design uses unprepared colonoscopy techniques to sample
each location of the gut without prior disruption of the native flora in
20 healthy volunteers. To get around the noise created by a diverse set
of human microbiomes, we used a machine-learning classification
algorithm trained on curated 16S rRNA sequencing reads to identify
microbes specific to each location. We found that our classification
models were able to separate mucosal and lumenal samples as well as
differentiate between sides of the colon based on populations of
specific microbes. By identifying the specific microbes we are poised to
ask if and how the presence or disruption of the microbes at each site
contribute to the development of the specific tumor subtypes of CRC in
the proximal and distal human colon.

@@

Proximal and distal lumenal and mucosal samples were collected from 20
healthy humans that had not undergone bowel preparation(Figure 1).
Participants also collected stool at home one week prior to the
procedure. To characterize the bacterial communities present at these
sites, 16S rRNA gene sequencing was performed on extracted DNA from each
sample. The relative abundance of each phyla for each site is shown in
Figure 2A. This data combines the samples from all participants in the
study. Each site is primarily dominated by \emph{Firmicutes} and
\emph{Bacteriodetes}, consistent with known variability in human
microbiome research. Likewise, samples had varying levels of diversity
at each site, irrespective of the individual (Figure 2B). That is, while
some individuals had a more diverse right mucosa, some had a more
diverse left mucosa, and so on. Therefore we could not identify a clear
pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (left or right) or sites (lumen or
mucosa), we calculated theta YC distances from OTU abundances and
compared distances for all individuals. Again, across all patient
samples we saw a range of theta YC distances when comparing sample
locations (Figure 3A) and again those ranges did not follow a clear
pattern on an individual basis. However, when comparing median distances
between the right lumen and mucosa, the left versus right lumen, the
left versus right mucosa and the left lumen and mucosa, we found that
the right lumen and mucosa were most similar to each other than the
other samples (P \textless{} 0.005, Wilcoxon, BH adjustment). Next, we
calculated thetaYC distances to examine how each sample compared to the
home-collected exit stool. Amidst variability between patients, we did
identify significantly smaller thetaYc distance between the left lumenal
sample and the exit stool (Figure 3B, P \textless{} 0.05, Wilcoxon, BH
adjustment). Further, there was an even larger difference in the
comparisons of the left mucosa to the exit stool, indicating that the
mucosa is quite different from the stool as compared to lumen (P
\textless{} 0.0005, Wilcoxon, BH adjustment). To determine what factors
may be driving the differences seen among the samples, we compared
thetaYC distances between samples from all subjects (interpersonal)
versus samples from within one subject (intrapersonal). We found that
samples from one individual were overall much more similar to each other
than to other study subjects (Figure 3C), consistent with previous human
microbiome studies that have sampled multiple sites of the human colon
(\textbf{???}). Thus interpersonal variation between subjects drives the
differences between samples more than sample site or location. Overall,
the results of the similarity analysis suggest that the contents of the
left lumen are most representative of stool at exit, and the microbes
remaining on the mucosa are much different.

@@

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest machine-learning models built on OTU
abundances. The first model we built to classify the lumen versus
mucosal samples for the right and left sides, independently (Figure 4A).
The constructed model used ((Xopt)) features for the right and ((Xopt))
for the left. The models performed well when classifying these samples
(AUC = 0.797 and AUC = 0.980, respectively). The OTUs that were most
predictive of each site are identified by their greatest mean decrease
in accuracy when removed from the model. For distinguishing the right
lumen and mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
were differentially abundant (Figure 4B). The model classifying the left
lumen and mucosa identified OTUs from \emph{Turicibacter},
\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
\emph{Anaerococcus} genera that could distinguish lumen from mucosal
samples (Figure 4C). These results indicate that there are fine
differences between the different sites of the colon, and that these can
be traced down to specific OTUs on each side.

@@

Given that specific bacterial species have been associated with
colorectal cancer we probed our sample set for these OTUs. Among our 100
samples, the most frequent sequence associated wtih the
\emph{Fusobacterium} genus was OTU179, which BLASTs to
\emph{Fusobacterium nucleatum subsp animalis}. It was the only sequence
to align with a \emph{F. nucleatum} strain- the only species of
\emph{Fusobacterium} known to have oncogenic properties and be found on
the surfaces of colorectal cancer tumors. ((12)). The
\emph{Fusobacterium} positive samples were located on the right and left
mucosa and ranged up to 1\% of the sample (Figure 6A). OTU00152 is
\emph{Porphyromonas} and the most frequent sequence in that OTU aligns
to \emph{Porphyromonas asacharolytica}, another bacterium commonly
detected and isolated from colorectal tumors. OTU152 was only detected
on the left mucosa, and in fact was one of the OTUs the classification
model identified as separating left and right sides (Figure 6B). Samples
that were positive for \emph{P. asacharolytica} ranged in abundances
from 0.01\% - 16\%. Thus, cancer-associated OTUs could be found in our
sample set of 20 healthy individuals.

@@

When examining the relative abundance of the different phyla at each
site, there is a wide amount of variation for each phyla with
communities primarily dominated by the Bacteriodes and Firmicutes. This
likely reflects not only the variability between human subjects, casued
by differences in age, gender, diet, but also reports of microbial
``patchiness'' in the gut microbiome. Several previous studies have
noted that the bacteria recoverable from the same mucosal sample
location can be vastly different when the samples are taken just 1 cm
away from each other ((13)). Similar patchiness is also observed in
lumenal contents and fecal samples themselves; there is observed
separation of different interacting microbes along the length of a stool
sample, for instance ((14)). That said, across our samples the mucosal
samples harbor more Proteobacteria, consistent with previous studies
comparing mucosal swabs to lumenal content in humans ((4)). Hence, the
conclusions we can draw from phyla analysis are likely confounded by
differences in sampling and patchiness between subjects.

We identified a range of Simpson diversity values between patients and
between each anatomical site. As this measurement takes both number of
species and abundance of species into account, it seems that there is a
varied level of each component along the human colon. There was no clear
pattern of diversity when comparing patients- some patients had a higher
level of diversity in the proximal than the distal colon, some patients
had the opposite. An early study of microbial diversity and membership
of three human subjects found a varied level of diversity in each
subject, but that subjects were overall more similar to themselves than
to each other (\textbf{???}). We used the thetaYC dissimilarity metric
to compare the communities present at each anatomical location to each
other. We found that while there was a wide range of similarity of each
site within a person, the study participants were ultimately more
similar to themselves than all of the other participants. This finding
is consistent with numerous studies examining the biogeography of the
colon in healthy humans (\textbf{???}, (15), (16)).

@@

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 7 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans ((9)) and have oncogenic
properties in cell culture and in mice ((17)). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the left mucosa. Not much is known
about the distribution of \emph{P. asacharolytica} but given its
documented anaerobic characteristics and asacharolytic metabolism, it
might not be surprising that it resides in the less-oygen-rich and
proteinaceous left mucosa ((4)). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
on right-sided tumors, and distribution of \emph{F. nucleatum} decreases
along the colon to rectum ((18)). Of the (8) (40\%) individuals positive
for \emph{F. nucleatum} in this study, the bacteria was spread across
the right mucosa, left lumen and left mucosa. That said, in a previous
study of 1,102 colorectal cancer cases, only 138 (11\%) of them
contained detectable \emph{F. nucleatum} at all ((18)). In a separate
study of patients with inflammatory bowel disease (IBD), 50\% of 22 IBD
patients and 17.6\% of 34 people without disease contained \emph{F.
nucleatum} in mucosal biopsies ((19)). These studies have the added
caveat of being sampled from prepared or cleaned colons. Thus, the
presence of \emph{F. nucleatum} in a healthy individual is not
necessarily linked to the development of future colorectal cancers.
Because of the spatial distribution of the \emph{F. nucleatum} in our
sample set, we cannot develop a model for the role of \emph{F.
nucleatum} in the healthy colon. Indeed, rather than the \emph{F.
nucleatum} initiating tumorigenesis, these bacteria may instead arrive
``late'' to the tumor microenvironment and feed on tumors that have
alread begun to develop ((20)). Data examining bacterial biofilms on CRC
tumors suggests that \emph{Fusobacteria} species are more commonly found
both on proximal tumors and in biofilms, indicating that it is not only
the presence of the bacteria but the organization of the tumor community
that contributes to \emph{Fusobacterium's} role in tumorigenesis.
Finally, \emph{Fusobacteria} and \emph{Porphyromonas} species have been
known to not only co-occur on CRC tumors but also to synergistically
promote tumorigenesis in an oral cancer model ((21), (20)). Thus,
further analysis of the distribution and role of these pathogens may
elucidate a mechanism for development of CRC tumors in the proximal or
distal colon.

Specific comparisons of our findings to previously published gut
biogeography studies are additionally confounded by the use of bowel
preparation methods in most other studies. A rare report of a
matched-colonoscopy study that sampled 18 patient's colonic mucosa and
lumenal contents prior to and after bowel cleansing ((22)). This group
found that mucosa and lumenal samples were distinguishable prior to
bowel cleansing, but that bowel preparation resulted in an increase in
shared OTUs between each site ((22)). Bowel cleansing not only made the
samples harder to distinguish, it resulted in decreases in diversity
across sites. Further, the differences were not great enough to overcome
interpersonal differences between subjects. So, bowel preparation
clearly induces bias into the microbes recovered from sampling the lumen
or mucosa of a prepared bowel. Thus our findings of specific bacteria at
each site of the colon are strengthened by the lack of bowel
preparation.

@@

\subsubsection{Figure legends}\label{figure-legends}

@@

@@

@@

@@

@@

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Specific taxa differentiate proximal and distal microbiota in the unprepped human colon},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{\emph{Specific taxa differentiate proximal and distal microbiota in the
unprepped human colon}}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide
despite improved preventative and therapeutic measures. Tumors of the
proximal (right) and distal (left) colon are morphologically and
genetically distinct. Previous work from our group found that microbial
dysbiosis is associated with the development of CRC tumors in studies of
both mice and humans. Analysis of the fecal microbiota from healthy and
CRC patients further revealed different microbial signatures associated
with disease. We extended our observations of the fecal microbiome to
include analysis of the proximal and distal human colon. We used a
two-colonoscope approach on subjects that had not undergone standard
bowel preparation procedure. This technique allowed us to characterize
the native proximal and distal luminal and mucosal microbiome without
prior chemical disruption. 16S rRNA gene sequencing was performed on
proximal and distal mucosal and luminal biopsies and home-collected
stool for 20 healthy individuals. Diversity analysis revealed that each
site contained a diverse community, and that a patient's samples were
more similar to each other than to that of other individuals. Since we
could not differentiate sites along the colon based on community
structure or community membership alone, we employed the Random Forest
machine-learning algorithm to identify key species that distinguish
biogeographical sites. Random Forest classification models were built
using taxa abundance and sample location and revealed distinct
populations that were found in each location. \emph{Peptoniphilus,
Anaerococcus, Enterobacteraceae, Pseudomonas} and \emph{Actinomyces}
were most likely to be found in mucosal samples versus luminal samples
(AUC = 0.925). The classification model performed well (AUC = 0.912)
when classifying mucosal samples into proximal or distal sides, but
separating luminal samples from each side proved more challenging (AUC =
0.755). The left mucosa was found to have high populations of
\emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
distal luminal samples were comprised of many of the same taxa, likely
reflecting the fact that stool moves along the colon from the proximal
to distal end. Finally, comparison of all samples to fecal samples taken
at exit uncovered that the feces were most similar to samples taken from
the left lumen, again reflecting the anatomical structure of the colon.
By sampling the unprepped human colon, our results have identified
distinct bacterial populations native to the proximal and distal sides.
Further investigation of these bacteria may elucidate if and how these
groups contribute to differential oncogenesis processes on the
respective sides of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related
deaths in the United States. CRC tumors vary in structure, size,
morbidity and symptomology depending upon their geographical location in
the colon. CRC has recently been described to fall into four molecular
subtypes ((1)). Of these subtypes, tumors that arise on the left side of
the colon are infiltrating lesions that present with painful symptoms.
In contrast, 47\% of CRCs are caused by right-sided colon tumors that
are bulky and project into the lumen, often remaining asymptomatic until
advanced carcinogenesis ((2)). Due to the absence of symptoms,
right-sided tumors have a significantly higher mortality rate ((2)). The
left and right sides of the colon differ in the amount of inflammation
present and the genomic instability of precancerous cells, respectively,
as well as oxygen, pH and the presence of antimicrobial peptides ((3),
(4), (5)). Microenvironments differ not only longitutinally along the
colon, but latitudinally from the epithelium to mucosa to intestinal
lumen, offering several sites for different microbial communities to
flourish. Given this varied physiology of the proximal-distal axis of
the colon, symbiotic microbes and their metabolites likely vary between
sites.

Several recent findings have shown that development and progression of
CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host ((3)). Our group
and others have found that the stool microbiome of patients with CRC is
distinct from that of healthy people ((6)). Further studies manipulating
the gut microbiome using antibiotics or other chemoagents in mice has
shown that dysbiosis preceded and accelerated the development of CRC
tumors ((7)). Comparison of the specific bacteria present on CRC tumors
with those found on nearby healthy tissue has also identified specific
bacterial species that are tumor-associated ((8)). These species include
the oral pathogens \emph{Fusobacterium nucleatum} and
\emph{Porphyromonas asacharolytica}. Interestingly, these periodontal
pathogens have been highly predictive of whether a patient had CRC
tumors or not in our classification studies ((9)). Because these studies
were performed in mice or examined only shed human stool, they were
unable to analyze paried samples from the proximal and distal sides of
the colon. Similarly, comparisons of on- or off-tumor bacteria rarely
have matched tissue from the other side of the colon from the same
patient, limiting what conclusions can be drawn about the colonic
microbiome overall, let alone at that specific site. Due to these
limitations, the contribution of the gut microbiota to these subtypes is
largely undefined. Characterizing these communities could provide needed
insight into CRC etiology, including how the disruption of the healthy
community could promote the initiation or proliferation of the distinct
left and right CRC tumors.

Further, the few existing profiles of the microbial biogeography of the
gut have been limited by sample collection methods. The majority of
human gut microbiome studies have been performed on whole shed stool or
on samples collected during colonoscopy procedures. While the latter
method allows investigators to acquire samples from inside the human
colon, typically this procedure is preceded by the use of bowel
preparation methods such as the consumption of laxatives to cleanse the
bowel. Bowel preparation is essential for detecting cancerous or
precancerous lesions in the colon, but complicates microbiome profiling
as the chemicals strip the bowel of contents and disrupt the mucosal
layer ((10), (11)). As such, what little information we do have about
the biogeographical distribution of the microbes in the proximal and
distal colon is confounded by the bowel preparation procedure.

We combined these approaches to address the limitations of previous
studies and effectively characterize the microbiome in the lumen and
mucosa of the proximal and distal healthy human colon. Our design used
unprepared colonoscopy techniques to sample each location of the gut
without prior disruption of the native bacteria in 20 healthy
volunteers. To address noise created by a diverse set of human
microbiomes, we used a machine-learning classification algorithm trained
on curated 16S rRNA sequencing reads to identify microbes specific to
each location. We found that our classification models were able to
separate mucosal and lumenal samples as well as differentiate between
sides of the colon based on populations of specific microbes. By
identifying the specific microbes we are poised to ask if and how the
presence or disruption of the microbes at each site contribute to the
development of the specific tumor subtypes of CRC in the proximal and
distal human colon.

@@

Proximal and distal lumenal and mucosal samples were collected from 20
healthy humans that had not undergone bowel preparation(Figure 1).
Participants also collected stool at home one week prior to the
procedure. To characterize the bacterial communities present at these
sites, 16S rRNA gene sequencing was performed on extracted DNA from each
sample. The relative abundance of each phyla for each site is shown in
Figure 2A. This data combines the samples from all participants in the
study. Each site is primarily dominated by \emph{Firmicutes} and
\emph{Bacteriodetes}, consistent with known variability in human
microbiome research (cite). Likewise, samples had varying levels of
diversity at each site, irrespective of the individual (Figure 2B). That
is, while some individuals had a more diverse right mucosa, some had a
more diverse left mucosa. Therefore we could not identify a clear
pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (left or right) or sites (lumen or
mucosa), we calculated theta YC distances from OTU abundances and
compared distances for all individuals. Again, across all patient
samples we observed a range of theta YC distances when comparing sample
locations (Figure 3A) and again those ranges did not follow a clear
pattern on an individual basis. However, when comparing median distances
between the right lumen and mucosa, the left versus right lumen, the
left versus right mucosa and the left lumen and mucosa, we found that
the right lumen and mucosa were most similar to each other than the
other samples (P \textless{} 0.005, Wilcoxon, BH adjustment).

Next, we calculated thetaYC distances to examine how each sample
compared to the home-collected exit stool. Amidst variability between
patients, we did identify significantly smaller thetaYc distance between
the left lumenal sample and the exit stool (Figure 3B, P \textless{}
0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger
difference in the comparisons of the left mucosa to the exit stool,
indicating that the mucosa is different from the stool as compared to
lumen (P \textless{} 0.0005, Wilcoxon, BH adjustment). To determine what
factors may be driving the differences seen among the samples, we
compared thetaYC distances between samples from all subjects
(interpersonal) versus samples from within one subject (intrapersonal).
We found that samples from one individual were overall much more similar
to each other than to other study subjects (Figure 3C), consistent with
previous human microbiome studies that have sampled multiple sites of
the human colon (\textbf{???}, (12), (13)). Thus interpersonal variation
between subjects drives the differences between samples more than sample
site or location. Overall, the results of the similarity analysis
suggest that the contents of the left lumen are most representative of
stool at exit, and the microbes remaining on the mucosa are much
different.

@@

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models built on OTU abundances. We
built the first model to classify the lumen versus mucosal samples for
the right and left sides, independently (Figure 4A). The constructed
model used ((Xopt)) features for the right and ((Xopt)) for the left.
The models performed well when classifying these samples (AUC = 0.797
and AUC = 0.980, respectively). The OTUs that were most predictive of
each site are identified by their greatest mean decrease in accuracy
when removed from the model. For distinguishing the right lumen and
mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
were differentially abundant (Figure 4B). The model classifying the left
lumen and mucosa identified OTUs from \emph{Turicibacter},
\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
\emph{Anaerococcus} genera that could distinguish lumen from mucosal
samples (Figure 4C). These results indicate that there are fine
differences between the different sites of the colon, and that these can
be traced down to specific OTUs on each side.

@@

Given that specific bacterial species have been associated with
colorectal cancer, we probed our sample set for these OTUs. Among our
100 samples, the most frequent sequence associated wtih the
\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
\emph{Fusobacterium nucleatum subsp animalis}. It was the only sequence
to align with a \emph{F. nucleatum} strain- the only species of
\emph{Fusobacterium} known to have oncogenic properties and be found on
the surfaces of colorectal cancer tumors. ((14)). The
\emph{Fusobacterium} positive samples were located on the right and left
mucosa and ranged up to 1\% of the sample (Figure 6A). OTU00152 is
\emph{Porphyromonas} and the most frequent sequence in that OTU aligns
to \emph{Porphyromonas asacharolytica}, another bacterium commonly
detected and isolated from colorectal tumors. OTU152 was only detected
on the left mucosa, and in fact was one of the OTUs the classification
model identified as separating left and right sides (Figure 6B). Samples
that were positive for \emph{P. asacharolytica} ranged in abundances
from 0.01\% - 16\%. Thus, cancer-associated OTUs could be found in our
sample set of 20 healthy individuals.

@@

When examining the relative abundance of the different phyla at each
site, there was a wide amount of variation for each phyla with
communities primarily dominated by the Bacteriodes and Firmicutes. This
likely reflects not only the variability between human subjects, casued
by differences in age, gender, diet, but also reports of microbial
``patchiness'' in the gut microbiome. Several previous studies have
noted that the bacteria recoverable from the same mucosal sample
location can be vastly different when the samples are taken just 1 cm
away from each other ((15)). Similar patchiness is also observed in
lumenal contents and fecal samples themselves; there is observed
separation of different interacting microbes along the length of a stool
sample, for instance ((16)). That said, across our samples the mucosal
samples harbor more Proteobacteria, consistent with previous studies
comparing mucosal swabs to lumenal content in humans ((4)). Hence, the
conclusions we can draw from phyla analysis are likely confounded by
differences in sampling and patchiness between subjects.

@@

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans ((9)) and have oncogenic
properties in cell culture and in mice ((17)). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the left mucosa. Not much is known
about the distribution of \emph{P. asacharolytica} but given its
documented anaerobic characteristics and asacharolytic metabolism, it
might not be surprising that it resides in the less-oygen-rich and
proteinaceous left mucosa ((4)). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
on right-sided tumors, and distribution of \emph{F. nucleatum} decreases
along the colon to rectum ((18)). Of the (8) (40\%) individuals positive
for \emph{F. nucleatum} in this study, the bacteria was spread across
the right mucosa, left lumen and left mucosa. The presence of \emph{F.
nucleatum} in a healthy individual is not necessarily linked to the
development of future colorectal cancers. Because of the spatial
distribution of the \emph{F. nucleatum} in our sample set, we cannot
develop a model for the role of \emph{F. nucleatum} in the healthy
colon. Data examining bacterial biofilms on CRC tumors suggests that
\emph{Fusobacteria} species are more commonly found both on proximal
tumors and in biofilms, indicating that it is not only the presence of
the bacteria but the organization of the tumor community that
contributes to \emph{Fusobacterium's} role in tumorigenesis ((8)).
Finally, \emph{Fusobacteria} and \emph{Porphyromonas} species have been
known to not only co-occur on CRC tumors but also to synergistically
promote tumorigenesis in an oral cancer model ((19), (20)). Thus,
further analysis of the distribution and activities of these pathogens
may elucidate a mechanism for development of CRC tumors in the proximal
or distal colon.

Specific comparisons of our findings to previously published gut
biogeography studies are additionally confounded by the use of bowel
preparation methods in most other studies. A rare report of a
matched-colonoscopy study that sampled 18 patient's colonic mucosa and
lumenal contents prior to and after bowel cleansing ((21)). This group
found that mucosa and lumenal samples were distinguishable prior to
bowel cleansing, but that bowel preparation resulted in an increase in
shared OTUs between each site ((21)). Bowel cleansing not only made the
samples harder to distinguish, it resulted in decreases in diversity
across sites. Further, the differences were not great enough to overcome
interpersonal differences between subjects. So, bowel preparation
clearly induces bias into the microbes recovered from sampling the lumen
or mucosa of a prepared bowel. Thus our findings of specific bacteria at
each site of the colon are strengthened by the lack of bowel
preparation.

@@

\subsubsection{Figures}\label{figures}

\includegraphics{submission/figure_1.pdf}

@@

\includegraphics{submission/figure_2.pdf}

@@

\includegraphics{submission/figure_3.pdf}

@@

\includegraphics{submission/figure_4.pdf}

@@

\includegraphics{submission/figure_5.pdf}

@@

\includegraphics{submission/figure_6.pdf}
